---
title: "fctutils: Advanced Factor Manipulation Utilities for R"
author:
- Kai Guo, University of Michigan
date: "`r Sys.Date()`"
output:
  html_document:
  df_print: paged
  word_document:
    toc: yes
    toc_depth: '6'
  rmarkdown::html_vignette: default
  pdf_document:
    toc: yes
    toc_depth: 6
vignette: |
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{fctutils}
  %\VignetteEngine{knitr::knitr}
---
The `fctutils` package provides a comprehensive suite of utilities for advanced manipulation and analysis of factor vectors in R. It offers tools for splitting, combining, reordering, filtering, and transforming factor levels based on various criteria. Designed to enhance the handling of categorical data, `fctutils` simplifies complex factor operations, making it easier to preprocess and analyze data in R.

**Key Features**:

- **Advanced Sorting and Reordering**: Sort factor levels based on custom functions, external vectors, or associated data.
- **Factor Level Manipulation**: Functions to split, collapse, map, and extract substrings from factor levels.
- **Grouping and Aggregation**: Aggregate factor levels based on grouping variables or hierarchical categories.
- **Handling Missing Data**: Impute missing values in factors using various methods.
- **Consistency Across Datasets**: Ensure consistent factor levels across multiple vectors or datasets.


##  1. Software Usage
### 1.1 Installation
Install the package with its dependencies and load it for usage in R.
``` {r install, eval = FALSE}
library(devtools) # Load the devtools package
install_github("guokai8/fctutils") # Install the package
```
## 2. Useful functions
### 2.1 Ordering and Sorting Factors
_fct_pos_ Reorders the levels of a factor vector based on the characters at specified positions within the factor levels.
```{r}
library(fctutils)
factor_vec <- factor(c('Apple', 'banana', 'Cherry', 'date', 'Fig', 'grape'))
# Reorder based on positions 1 and 3, case-insensitive
fct_pos(factor_vec, positions = c(1, 3))
# Reorder in decreasing order, case-sensitive
fct_pos(factor_vec, positions = 1:2, case = TRUE, decreasing = TRUE)
```

_fct_count_ Reorders the levels of a factor vector based on the count of each level in the data.
```{r}
factor_vec <- factor(c('apple', 'banana', 'apple', 'cherry', 'banana', 'banana', 'date'))

# Reorder levels by decreasing count
fct_count(factor_vec)
# Reorder levels by increasing count
fct_count(factor_vec, decreasing = FALSE)
```
_fct_sub_ Reorders the levels of a factor vector based on substrings extracted from the factor levels.
```{r}
factor_vec <- factor(c('Apple', 'banana', 'Cherry', 'date', 'Fig', 'grape'))
# Reorder based on substring from position 2 to 4
fct_sub(factor_vec, start_pos = 2, end_pos = 4)

# Reorder from position 3 to end, case-sensitive
fct_sub(factor_vec, start_pos = 3, case = TRUE)
```
_fct_freq_ Reorders the levels of a factor vector based on the total frequency of characters appearing in the vector.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date', 'banana', 'apple', 'fig'))

# Reorder levels based on total character frequency
fct_freq(factor_vec)

# Reorder levels, case-sensitive
factor_vec_case <- factor(c('Apple', 'banana', 'Cherry', 'date', 'banana', 'apple', 'Fig'))
fct_freq(factor_vec_case, case = TRUE)
```
_fct_char_freq_ Reorders the levels of a factor vector based on the frequency of characters at specified positions within the data.
```{r}
factor_vec <- factor(c('apple', 'banana', 'apricot', 'cherry', 'banana', 'banana', 'date'))

# Reorder based on characters at positions 1 and 2
fct_char_freq(factor_vec, positions = 1:2)

# Reorder, case-sensitive, decreasing order
fct_char_freq(factor_vec, positions = c(1, 3), case = TRUE)
```
_fct_substr_freq_ Reorders the levels of a factor vector based on the frequency of substrings extracted from the data.
```{r}
factor_vec <- factor(c('苹果', '香蕉', '苹果', '樱桃', '香蕉', '香蕉', '日期'))

# Reorder based on substrings from position 1 to 1
fct_substr_freq(factor_vec, start_pos = 1, end_pos = 1)

# Reorder from position 2 to end
fct_substr_freq(factor_vec, start_pos = 2)
factor_vec <- factor(c('apple', 'banana', 'apricot', 'cherry', 'banana', 'banana', 'date'))
fct_substr_freq(factor_vec, start_pos = 2, end_pos=3)
```
_fct_regex_freq_ Reorders the levels of a factor vector based on the frequency of substrings matching a regular expression.
```{r}
factor_vec <- factor(c('apple', 'banana', 'apricot', 'cherry', 'blueberry', 'blackberry', 'date'))

# Reorder based on pattern matching 'a'
fct_regex_freq(factor_vec, pattern = 'a')

# Reorder with case-sensitive matching
fct_regex_freq(factor_vec, pattern = '^[A-Z]', case = TRUE)
```
_fct_split_ Splits the levels of a factor vector using specified patterns or positions and reorders based on specified parts or criteria.
```{r}
# Example factor vector with patterns
factor_vec <- factor(c('item1-sub1', 'item2_sub2', 'item3|sub3', 'item1-sub4'))

# Split by patterns '-', '_', or '|' and reorder based on the first part
fct_split(factor_vec, split_pattern = c('-', '_', '\\|'), part = 1)

# Use the second pattern '_' for splitting
fct_split(factor_vec, split_pattern = c('-', '_', '\\|'), use_pattern = 2, part = 2)

# Reorder based on character frequencies in the specified part
fct_split(factor_vec, split_pattern = '-', part = 2, char_freq = TRUE)

```

_fct_len_ Reorders the levels of a factor vector based on the character length of each level.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date'))

# Sort levels by length
fct_len(factor_vec)
```
_fct_sort_ Sorts the levels of a factor vector based on the values of another vector or a column from a data frame. Handles cases where the sorting vector may contain `NA`s.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date'))
by_vec <- c(2, 3, 1, NA)
fct_sort(factor_vec, by = by_vec)

# Example using a data frame column
data <- data.frame(
  Category = factor(c('apple', 'banana', 'cherry', 'date')),
  Value = c(2, 3, 1, NA)
)
fct_sort(data$Category, by = data$Value)
```
_fct_sort_custom_ Reorders the levels of a factor vector based on a custom function applied to each level.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry'))

# Sort levels by reverse alphabetical order
fct_sort_custom(factor_vec, function(x) -rank(x))

# Sort levels by length of the level name
fct_sort_custom(factor_vec, function(x) nchar(x))
```

### 2.2 Replacing Factor Levels
_fct_replace_ Replaces a specified level in a factor vector with a new level. If a position is provided, the new level is inserted at the specified position among the levels; otherwise, the original level order is preserved.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date', 'fig', 'grape'))

# replace 'banana' as 'blueberry', and keep original order
fct_replace(factor_vec, old_level = 'banana', new_level = 'blueberry')

# replace 'banana' as 'blueberry'
fct_replace(factor_vec, old_level = 'banana', new_level = 'blueberry', position = 2)
```
_fct_replace_pattern_ Replaces parts of the factor levels that match a specified pattern with a new string.
```{r}
factor_vec <- factor(c('apple_pie', 'banana_bread', 'cherry_cake'))

# Replace '_pie', '_bread', '_cake' with '_dessert'
fct_replace_pattern(factor_vec, pattern = '_.*', replacement = '_dessert')
```

### 2.3 Filtering and Removing Factor Levels
_fct_filter_freq_ Filters out factor levels that occur less than a specified frequency threshold and recalculates character frequencies excluding the removed levels. Offers options to handle NA values and returns additional information.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date', 'banana', 'apple', 'fig', NA))

# Filter levels occurring less than 2 times and reorder by character frequency
fct_filter_freq(factor_vec, min_freq = 2)

# Filter levels, remove NA values, and return additional information
result <- fct_filter_freq(factor_vec, min_freq = 2, na.rm = TRUE, return_info = TRUE)
result$filtered_factor
result$removed_levels
result$char_freq_table
```
_fct_filter_pos_ Removes factor levels where a specified character appears at specified positions within the levels.
```{r}
factor_vec <- factor(c('apple', 'banana', 'apricot', 'cherry', 'date', 'fig', 'grape'))

# Remove levels where 'a' appears at position 1
fct_filter_pos(factor_vec, positions = 1, char = 'a')

# Remove levels where 'e' appears at positions 2 or 3
fct_filter_pos(factor_vec, positions = c(2, 3), char = 'e')

# Case-sensitive removal
factor_vec_case <- factor(c('Apple', 'banana', 'Apricot', 'Cherry', 'Date', 'Fig', 'grape'))
fct_filter_pos(factor_vec_case, positions = 1, char = 'A', case = TRUE)
```
_fct_remove_levels_ Removes specified levels from a factor vector, keeping the remaining levels and their order unchanged.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date', 'fig', 'grape'))

# Remove levels 'banana' and 'date'
fct_remove_levels(factor_vec, levels_to_remove = c('banana', 'date'))
```
_fct_filter_func_ Removes levels from a factor vector based on a user-defined  function.
```{r}
factor_vec <- factor(c('apple', 'banana', 'cherry', 'date'))

# Remove levels that start with 'b'
fct_filter_func(factor_vec, function(x) !grepl('^b', x))
```
### 2.4 Merging Factor Vectors
_fct_merge_similar_ Merges levels of a factor that are similar based on string distance.
```{r}
factor_vec <- factor(c('apple', 'appel', 'banana', 'bananna', 'cherry'))

# Merge similar levels
fct_merge_similar(factor_vec, max_distance = 1)
```
_fct_concat_ Combines multiple factor vectors into a single factor, unifying the levels.
```{r}
factor_vec1 <- factor(c('apple', 'banana'))
factor_vec2 <- factor(c('cherry', 'date'))

# Concatenate factors
concatenated_factor <- fct_concat(factor_vec1, factor_vec2)
levels(concatenated_factor)
```
_fct_combine_ Combines two vectors, which may be of unequal lengths, into a factor vector and sorts based on the levels of either the first or second vector.
```{r}
vector1 <- c('apple', 'banana', 'cherry')
vector2 <- c('date', 'fig', 'grape', 'honeydew')

# Combine and sort based on vector1 levels
fct_combine(vector1, vector2, sort_by = 1)

# Combine and sort based on vector2 levels
fct_combine(vector1, vector2, sort_by = 2)
```
### 2.5 Other Useful Functions
_fct_intersect_ Combines multiple factor vectors and returns a factor vector containing only the levels common to all.
```{r}
factor_vec1 <- factor(c('apple', 'banana', 'cherry'))
factor_vec2 <- factor(c('banana', 'date', 'cherry'))
factor_vec3 <- factor(c('banana', 'cherry', 'fig'))

# Get intersection of levels
fct_intersect(factor_vec1, factor_vec2, factor_vec3)
```
_fct_union_ Combines multiple factor vectors and returns a factor vector containing all unique levels.
```{r}
factor_vec1 <- factor(c('apple', 'banana'))
factor_vec2 <- factor(c('banana', 'cherry'))
factor_vec3 <- factor(c('date', 'fig'))

# Get union of levels
fct_union(factor_vec1, factor_vec2, factor_vec3)
```
_fct_reorder_within_ Reorders the levels of a factor vector within groups defined by another factor vector.
```{r}
data <- data.frame(
  item = factor(c('A', 'B', 'C', 'D', 'E', 'F')),
  group = factor(c('G1', 'G1', 'G1', 'G2', 'G2', 'G2')),
  value = c(10, 15, 5, 20, 25, 15)
)
data <- rbind(data,data)

# Reorder 'item' within 'group' by 'value'
data$item <- fct_reorder_within(data$item, data$group, data$value, mean)
```
_fct_extract_ Extracts substrings from the levels of a factor vector based on a regular expression pattern and creates a new factor.
```{r}
factor_vec <- factor(c('item123', 'item456', 'item789'))

# Extract numeric part
fct_extract(factor_vec, pattern = '\\d+')

# Extract with capturing group
factor_vec <- factor(c('apple: red', 'banana: yellow', 'cherry: red'))
fct_extract(factor_vec, pattern = '^(\\w+):', capture_group = 1)
```
_fct_pad_levels_ Pads the levels of a factor vector with leading characters to achieve a specified width.
```{r}
# Example factor vector
factor_vec <- factor(c('A', 'B', 'C', 'D'))

# Pad levels to width 4 using '0' as padding character
padded_factor <- fct_pad_levels(factor_vec, width = 4, pad_char = '0')
print(levels(padded_factor))
# Output: "000A" "000B" "000C" "000D"

# Pad levels to width 6 using '%A' as padding string
padded_factor <- fct_pad_levels(factor_vec, width = 6, pad_char = '%A')
print(levels(padded_factor))
# Output: "%%A%A" "%%A%B" "%%A%C" "%%A%D"
```
_fct_level_stats_ Computes statistical summaries for each level of a factor vector based on associated numeric data. (group_by and summarize). 

_fct_pattern_replace_ Replaces substrings in factor levels that match a pattern with a replacement string. 

_fct_impute_ Replaces \code{NA} values in a factor vector using specified imputation methods. 

_fct_unique_comb_ Generates a new factor where each level represents a unique combination of levels from the input factors. 

_fct_map_func_ Transforms factor levels by applying a function that can include complex logic. 

_fct_collapse_lev_ Collapses specified levels of a factor into new levels based on a grouping list. 

_fct_duplicates_ Identifies duplicate levels in a factor vector and returns a logical vector indicating which elements are duplicates. 

_fct_dummy_ Generates a data frame of dummy variables (one-hot encoded) from a factor vector. 

_fct_replace_na_ Replaces \code{NA} values in a factor vector with a specified level. 

_fct_sample_levels_ Randomly selects a specified number of levels from a factor vector. 

_fct_apply_ Transforms factor levels by applying a function to each level. 

_fct_encode_ Converts the levels of a factor vector into numeric codes, optionally using a provided mapping. 

## 3. Summary
The `fctutils` package provides a comprehensive set of functions to efficiently manage and manipulate factor vectors in R. From ordering and sorting to replacing, filtering, merging, and beyond, these tools enhance your ability to handle categorical data with ease. The additional essential functions further extend the package's capabilities, ensuring that all common factor operations are covered.

## 4. Contact information
For any questions please contact guokai8@gmail.com or submit the issues to https://github.com/guokai8/fctutils/issues

